---
title: "DBI Update Table"
author: "Bruce C. Schardt"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DBI Update Table}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setupUT, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### S4 generic dbUpdateTable()

This DBI Extention package introduces a new S4 generic `dbUpdateTable`. Updating tables is not the focus of the DBI package as pure analysis scripts treat a database tables as read-only. 

In other domains, database maintainence requires updating information in rows of an existing table. Dropping the table and recreating it as is done by dbWriteTable with the `overwrite` argument set to `TRUE` is not an optimal solution for many applications.

No single generic update function could cover the  diverse variety of possible UPDATE queries. The `dbUpdateTable` S4 generic introduced in this package performs a specific type of DBMS UPDATE operation using the update join syntax. An example of this syntax is shown in the sql fragment shown below. This syntax is used by MariaDB and MySQL.

```{sql sqlUpdateTable, echo=TRUE, eval=FALSE}
UPDATE T1
INNER JOIN T2 ON T1.C1 = T2.C1 [AND T1.C2 = T2.C2]
SET T1.CX = T2.CX,
    T1.CY = T2.CY;
```

reference: [tutorial on UPDATE join syntax for MySQL](http://www.mysqltutorial.org/mysql-update-join/)

Note that no `WHERE` clause is specified for the query. This by design; the reasoning for which will be explained later in the document.

For `dbUpdateTable()` the table `T1` must be a pre-existing table in the DBMS. Table `T2` will be created as a temporary table in the DBMS using a dataset supplied as a function argument. This dataset will usually be a tibble, a data.frame, or another object coercible to a data.frame.

the `dbUpdateTable` function call has two additional arguements `set` and `by` used for constructing the `SET` and `ON` clauses of the UPDATE query.

```{r dbUpdateTable, echo=TRUE, eval=FALSE}
dbUpdateTable(con, "T1", T2, set = c("CX","CY"), by = c("C1","C2"))
```

### SQL Helper Functions

There is a helper function sqlUpdateTable which generates the UPDATE sql but does not execute it. Helper functions are also provided to generate the sql for the `ON` clause of the join as well as the multiples SET clauses of the `UPDATE`. These are `sqlOnClause()` and `sqlSetClause()`. 

```{r,echo=TRUE, eval=FALSE}
T2 <- tibble::as_tibble(list(C1=1:26,C2=LETTERS,CX=0,CY=1))
sqlUpdateTable(con, "T1", T2, set = c("CX","CY"), by = c("C1","C2"))
sqlOnClause(con,"T1","T2",by = c("C1","C2"));
sqlSetClause(con,"T1","T2",set = c("CX","CY"));
```

### Table Indexing

It is best practice that the target DBMS table be indexed, with the indexed columns matching the columns specified by the `by` argument. The temporary table used store the update dataset is indexed by the columns in the `by` argument as part of the `dbUpdateTable()` implementation. The `UPDATE-JOIN` operation with indexed table is generally quite fast.  

### Why no `WHERE` caluse?

Update queries can use a `WHERE` clause to limit the scope of the update. However in an `UPDATE-JOIN` the scope of the query is alreadly limited by the `ON` clause of the join. This combined with the power of `R` to precisely form the update dataset makes use of a `WHERE` clause redundent, IMHO.





