---
title: "DBI Update Table"
author: "Bruce C. Schardt"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DBI Update Table}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setupUT, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This DBI Extention package introduces a new S4 generic `dbUpdateTable`. Updating tables is not the focus of the DBI package as pure analysis scripts treat database tables as read-only. 

In other domains, database maintainence requires updating information in rows of an existing table. Dropping the table and recreating it as is done by dbWriteTable with overwrite argument set to true is not an optimal solution for many applications.

No single generic update function could cover the  diverse variety possible UPDATE queries. The `dbUpdateTable` S4 generic is introduced to perform a specific type  DBMS UPDATE operation using the update join syntax. An example of this syntax is shown in the sql fragment shown below

```{sql sqlUpdateTable, echo=TRUE, eval=FALSE}
UPDATE T1
INNER JOIN T2 ON T1.C1 = T2.C1 [AND T1.C2 = T2.C2]
SET T1.CX = T2.CX,
    T1.CY = T2.CY;
```

reference: [tutorial on UPDATE join syntax for MySQL](http://www.mysqltutorial.org/mysql-update-join/)

Note that no `WHERE` clause is specified for the query. This by design, the reasoning for which will be explained later in the document.

For `dbUpdateTable()` the table `T1` must be a pre-existing table in the DBMS. Table `T2` will be created as a temporary table in the DBMS using a dataset supplied as a function argument. This dataset will usually be a tibble, a data.frame, or another object coercible to a data.frame.

the `dbUpdateTable` function call has two additional arguements `set` and `by` used for constructing the `SET` and `ON` clauses of the UPDATE query.

```{r dbUpdateTable, echo=TRUE, eval=FALSE}
dbUpdateTable(con, "T1", T2, set = c("CX","CY"), by = c("C1","C2"))
```

There is a helper function sqlUpdateTable which generates the UPDATE sql but dowes not execute it.

```{r,echo=TRUE, eval=FALSE}
T2 <- tibble::as_tibble(list(C1=1:26,C2=LETTERS,CX=0,CY=1))
sqlUpdateTable(con, "T1", T2, set = c("CX","CY"), by = c("C1","C2"))
```

### Table Indexing

It is best practice to that the target DBMS table be indexed, with the indexed columns matching the columns specified by the `by` argument. When the temporary table used store the update dataset will be indexed by the columns in the `by` argument. The UPDATE-join operation with indexed table is generally quite fast.  

### Why no `WHERE` caluse?

Update queries can use a `WHERE` clause to limit the scope of the update. However in an Update-join the scope of the query is alreadly limited by the `ON` clause of the join. This combined with the power of `R` to precisely form the update dataset makes use of a `WHERE` clause redundent.





